Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{purity} \(p = 1\) if \textit{pure}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
  for this judgement are given in Figure \ref{fig:ssa-term-typing}; in
  particular, we note that a impure terms cannot be \textit{nested}; that is,
  all proper subterms of any term are pure.
  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and branches only to labels in
  \(\ms{L}\). The rules for this judgement are given in Figure
  \ref{fig:ssa-block-typing}.
  \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

@@

\TODO{some text about each typing rule}

>>>>

Our grammar proper is split into three main syntactic categories:
\begin{itemize}
  \item \textit{Expressions} \(a, b, c, d, e\), which are typed via the
  judgement \(\hasty{\Gamma}{p}{a}{A}{q}\), which says that the expression \(a\)
  is of type \(A\) with \textit{purity} \(p = 1\) if \textit{pure}, or \(p = 0\)
  otherwise, and \textit{quantity} \(q \subseteq \{\ms{a}, \ms{r}\}\). The rules
  for this judgement are given in Figure \ref{fig:ssa-term-typing}.
  \item \textit{Blocks} \(s, t\), which are typed via the judgement
  \(\haslb{\ms{L}}{t}{\Gamma}\), which says that the block \(t\) is well-formed
  with respect to the context \(\Gamma\) and branches only to labels in
  \(\ms{L}\). The rules for this judgement are given in Figure
  \ref{fig:ssa-block-typing}.
  \item \textit{Label-sets} or \textit{control-flow graphs} \(L\), which are
  typed via the judgement \(\lhaslb{\ms{L}}{L}{\ms{K}}\), which says that \(L\)
  defines a control-flow graph with inputs \(\ms{L}\) and outputs \(\ms{K}\)
  (some of which may also be in \(\ms{L}\), indicating recursive control-flow).
  The rules for this judgement are given in Figure \ref{fig:ssa-block-typing}.
\end{itemize}
Here \textit{contexts} \(\Gamma\), consist of a list of variables \(x\) given a
type \(A\) annotated with a \textit{quantity} \(q \subseteq \tint\). A variable
\(x: A^q\) is called \textit{affine} if its type \(A\) is affine (i.e.
\(\tlin{A}{\taff}\)) \textit{and} \(\taff \subseteq q\), in which case it can be
freely ignored. Similarly, \(x\) is \textit{relevant} if its type \(A\) relevant
(i.e. \(\tlin{A}{\trel}\)) \textit{and} \(\trel \subseteq q\), in which case it
can be freely ignored. This is formalized in the structural judgement
\(\csplits{\Gamma}{\Delta}{\Xi}\), which says that \(\Gamma\) can be split into
subcontexts \(\Delta\) and \(\Xi\) (weakening is simply the special case
\(\cwk{\Gamma}{\Delta} \iff \csplits{\Gamma}{\Delta}{\cdot}\)), as per the
typing rules in Figure \ref{fig:ssa-structural}. In particular, the rule
\rle{split-nil} provides a base case by saying that the empty context can be
split into two empty contexts, while the rules \rle{split-left} and
\rle{split-right} send \(x: A^q\) to the left or right subcontext repsectively;
both these rules also allow us to reduce the quantity \(q\) to a quantity \(r
\subseteq q\), "forgetting" some of our capabilities. The rule \rle{split-dup}
allows the duplication of a variable which has a relevant type \(A\)
\textit{and} a relevant quantity \(q\), with the option to weaken the quantity
to \textit{different} quantities \(r\) and \(s\) on the left-hand side and
right-hand sides respectively. Finally, the rule \rle{split-aff} simply allows
us to ignore a variable \(x: A^q\) if \(A\) and \(q\) are affine.

@@

We can now get to the typing judgement for expressions in Figure
\ref{fig:ssa-term-typing}. The rule \rle{var} is essentially the standard rule
for variables, with the requirement that, other than \(x\), all variables in
\(\Gamma\) are affine. Note that the quantity \(q\) of the \textit{expression}
\(x\) is must satisfy \(q \subseteq q'\), where \(x: A^{q'} \in \Gamma\); this
reflects that we're allowed to \textit{forget} capabilities, but not
\textit{add} them. The \rle{app} rule is also mostly standard: given an
instruction \(f\) which takes in an argument \(A\) and has a return type of
\(B\) and a \textit{pure} expression \(\hasty{\Gamma}{1}{a}{A}{q}\), we conclude
that \(\hasty{\Gamma}{p}{f\;a}{B}{q}\). We, of course, require that \(f\) itself
have purity \(p\) and quantity \(q\), i.e. that \(f \in \mc{I}_p^q(A, B)\). The
\rle{pair} rule says that we can type a pair \((a, b)\) in \(\Gamma\) if
\(\csplits{\Gamma}{\Delta}{\Xi}\) where \(\Delta\) types \(a\) and \(\Xi\) types
\(b\); the quantity of a pair is simply the intersection of the quantity of its
components (this is expressed in the rule by allowing any quantity which types
\textit{both} components), and both components must be pure. Finally, we have
the rules \rle{let} and \rle{let2}, corresponding to let-bindings and
destructuring let-bindings respectively. In both cases, we split the context
\(\Gamma\) into subcontext \(\Delta, \Xi\), one of which is used to type a pure
subterm \(a\), which is bound to variables \(x\) or destructured to variables
\(x, y\) respectively, in a potentially impure term \(e\) which is otherwise
typed in \(\Delta\).

@@
